---
title: "Assignment 2: Coding Basics"
author: Eric Newton
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq(1,30,3) #sequence starting at 1 to 30, by 3

sequencebythree <- seq(1,30,3) #assigning above sequence with descriptive name

#2. 
mean(sequencebythree)

median(sequencebythree) #using functions mean() and median() to calculate values
#3. 
mean(sequencebythree) > median(sequencebythree) #used greater than symbol to ask R

```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.



6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5.
student_names <- c("Peter", "Paul", "Luke", "John")
student_grades <- c(20, 40, 60, 80)
passing_grade <- c(FALSE, FALSE, TRUE, TRUE)

#6.
#student_names is a character vector; student_grades is a numerical vector; passing_grade is a logical vector

#7.
cbind(student_names, student_grades, passing_grade)
df_student_performance <- cbind(student_names, student_grades, passing_grade)


#8.
#columns already listed with informative names?


```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 


11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#10.

#11.
greater_than_50 <- ifelse(student_grades > 50, "TRUE", "FALSE")

print(greater_than_50)


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
