---
title: "Assignment 3: Data Exploration"
author: "Eric Newton"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
library(tidyverse)
library(lubridate)

Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:
Neonicotinoids are a class of pesticides widely used in global agriculture. It use is harmful to insects for a number or reasons. It does not specifically target pest insects, rather it affects all species of insects that come into contact with the chemical. It enters the plant itself when applied, so all parts of the plant incluiding pollen and nectar are toxic to insects including pollinator species. And, it remains in the soil after application and can travel signficant distances as runoff and in groundwater, affecting a far greater area than where it was applied.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:
Litter and woody debris serve as habitat for terrestrial and aquatic speices, provide soil stability and erosion prevention, and pool nutrients. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Elevated PVC litter trap design - elevated and ground-based traps that collect litter and woody debris
 2."Sampling is executed at terrestrial NEON sites that contain woody vegetation >2m
tall"
 3. "Trap placement within plots may be either targeted or randomized, depending on the vegetation."

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics) #dim provides dimensions of the dataset, by rows and column

```
> Answer:
The Neonics dataset contains 4,623 rows and 30 columns.

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect) #summary function for column with character values provides a frequency table with value name and its count in the column

```

> Answer:
The three most common effects that are studied are Population, Mortality, and Feeding Behavior.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}
summary(Neonics$Species.Common.Name) #summary by column again provides frequency data. Data already sorted by highest frequecy.
```

> Answer:
1. Honey Bee
2. Parasitic Wasp
3. Buff Tailed Bumblebee
4. Carniolan Honey Bee
5. Bumble Bee
6. Italian Honeybee
    These species are all essential for agriculture. The bee species are all important pollinators and parasitic wasps control the populations of pests species. 


8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}
class(Neonics$Conc.1..Author.) #class() shows the class of the column you enter 
```

> Answer:The class function shows that Conc.1..Author is a factor because we used the subcommand to read strings as factors. This is a character vector. We know it is not numeric becuase the vector non-numeric values, such as 12/.


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year)) #provides frequency plot of studies by publication year, x values defined as Publication.Year column

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location, binwidth = 20)) #added color = Test.Location to make each test location distinct on the plot
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:
The most Common

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}

```

> Answer:

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate) #class() provides info on class of a variable

ymd(Litter$collectDate) #assigned 'collectDate' as date class, ymd because that is the order in the data

August_2018 <- unique(Litter$collectDate, incomperables = FALSE) 
August_2018

```

> Answer:
Litter was sampled on August 2nd and 30th.

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
Niwot_Ridge <- unique(Litter$plotID, incomparables = FALSE)
Niwot_Ridge
```

> Answer:

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
ggplot(Litter, aes(x = functionalGroup)) + geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
