---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Eric Newton"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()

library(tidyverse)
library(agricolae)
library(lubridate)
library(here)

Nutrients <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
                      stringsAsFactors = TRUE)

Nutrients$sampledate <- mdy(Nutrients$sampledate)

#read in dataset and coverted sampledate to a date vector

#2
mytheme <- theme_classic(base_size = 14)+
  theme(legend.background = element_rect(
    color = "grey",
    fill = "white"),
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom")

#created theme and set to mytheme

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded during July does not change with depth 
across all lakes.
Ha: Mean lake temperature recorded during July does change with depth across
all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
Nutrients_processed <- 
  Nutrients %>%
  filter(month(sampledate) == 07) %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  na.omit(Nutrients)

#wrangled data by filtering for temperatures recorded in July, selecting
#for the vectors of interest and ommiting NAs

#5

tempbydepth <- ggplot(Nutrients_processed, aes(
  x = depth,
  y = temperature_C)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "orange") +
  labs(
    title = "Effect of Depth on Lake Temperature",
    x = "Lake Depth (meters)",
    y = "Temperature (C)"
  ) +
  guides(color = guide_legend(title = NULL)) +
  mytheme
  ylim(0, 35)

print(tempbydepth)

#created scatter plot with a linear regression line using geom_point and 
#geom_smooth and changed color, labels, and legend to make it easier to read
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
The figure suggests that temperature decreases with increasing depth. Temperature appears to fall somewhat linearly at low depths but flattens to a temperature around 7 or 8C at lower depths. 

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
tempbydepth_regression <- lm(data = Nutrients_processed,
                           temperature_C ~ depth)
summary(tempbydepth_regression)

#used lm function to perform a linear regression for the relationship between
#temperature and depth
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:
Approximately 74% of variablity in lake temperature can be explained by 
changes in depth. There are 9,726 degrees of freedom, meaning that there are 
9,727 levels in the data. The p-value is less than 0.05, so that the null 
hypothesis can be rejected and the findings are significant. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
tempAIC <- lm(data = Nutrients_processed,
              temperature_C ~ year4 + daynum + depth)
step(tempAIC)
summary(tempAIC)

#10
temp_multiple <- lm(data = Nutrients_processed,
              temperature_C ~ year4 + daynum + depth)

#ran an AIC using step function to assess which variables most explain variance
#in temperature. Then ran a multiple analysis with the variables outputted by
#the AIC analysis. 
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The AIC method suggested that we use all three explanatory variables - year, day number, and depth. The model explains 74.1% of the observed variance. There is a slight improvement, as the R-squred value increased by ~0.3%. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
tempbylake_anova1 <- aov(data = Nutrients_processed,
                        temperature_C ~ lakename)
summary(tempbylake_anova1)

tempbylake_anova2 <- lm(data = Nutrients_processed,
                        temperature_C ~ lakename)
summary(tempbylake_anova2)

#perfomed an anova test uisng aov and lm functions and summarized the tests to 
#mean estimates and p-values

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
There are significant differences in mean temperature among the lakes, including a 7 degree difference bewteen Central Long Lake and East Long Lake. All values had a p-value less than 0.05. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
tempbydepth2 <- 
ggplot(Nutrients_processed, aes(
  x = depth, 
  y = temperature_C, 
  color = lakename, 
  group = lakename)) +
geom_point(alpha = 0.5) +
geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Depth and Temperature Across Lakes in North Temperate Lakes LTER", 
       x = "Lake Depth (meter)", 
       y = "Temperature (C)") +
  guides(color = guide_legend(title = NULL)) +
  mytheme +
  ylim(0, 35)

print(tempbydepth2)

#created a scatterplot with a linear regression line using ggplot and 
#selected color and group as 'lakename' in aesthetics to differentiate plot
#points and lm line by lakename
```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(tempbylake_anova1)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Lakes with the same mean temperature based on the p-values in the Tukey HSD test
are Paul Lake and Ward Lake. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 
We can use a two-sample t-test to assess the difference in means between two
variables, such as Peter Lake and Paul Lake. 


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
CramptonandWard <- 
  Nutrients_processed %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

temp_t.test <- t.test(CramptonandWard$temperature_C ~ CramptonandWard$lakename)
temp_t.test

#wrangled the processed dataset to filter for Crampton and Ward Lakes, then
#ran a two-sample t-test to determine if the mean temperature in each lake
#is equal
```

>Answer: 
The two-sample T-test showed that the means are the same. The null hypothesis
is accepted, as the p-value is greater than 0.05. That does match the results of
the Tukey HSD test, which showed that Crampton and Ward lake have the same
means, statistically speaking. 